Crime in Austin Texas: Exploration, Classification, and Trends

### 2024-06-22 code/eda-code/eda.qmd
Lots of exploratory work done, deep dive into time series data, and many of the categorical variables in eda code. Moved a lot of that work out of the manuscript. Added descriptions, ideas, and interpretations. Started working on an ARIMA model for crime frequency.

version https://git-lfs.github.com/spec/v1
oid sha256:a063832bfc7ed6b54f7db32a75d484bdb6f59df9d1fba1d7ca6f90fe7911162f
size 4505

>>>>

---
title: "An example exploratory analysis script"
date: "2024-02-07"
output: html_document
---

# Setup

```{r setup}
#load needed packages. make sure they are installed.
pacman::p_load(here, knitr, tidyverse, skimr, fpp2)
theme_set(theme_minimal())
```

Load the data.

```{r}
d1 <- readRDS(here('data','processed-data','processed-crime.rds'))
zips <- read_csv(here('data','raw-data','austin-zip-codes.csv'))
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary(d1)
head(d1)
str(d1)
skim(d1)
```

```{r}
summary(zips)
head(zips)
str(zips)
skim(zips)
```

Important to note that the dataset has several date fields, and that alone will be ripe with opportunity for exploration. However I want to take a look at some of the other variables first, to see if maybe isolating to one of the other categories, or including other categorical variables in the time series could be useful. First off is Zip Code, because location is often a major factor in determining the amount and type of crime that would occur.

```{r}
d1 %>% 
  mutate(years = year(Occurred.Date),
         years = as.character(years)
         ) %>% 
  group_by(years, Zip.Code) %>% 
  summarize(
    count = n()
  ) %>% as.data.frame() %>% pivot_wider(names_from = years, values_from = count)
```

Layering in the year gives an interesting view, because it gives us the opportunity to see major increases in certain locations over time. However, there are a lot of zip Codes and a lot of years in this dataset. A visualization may help with this, my first thought being a chloropeth plot with a slider for the year value. Other quicker solutions might be looking at the individual years as observations of sorts, then looking at summary statistics. More to come here.

```{r}
d1 %>% 
  mutate(years = year(Occurred.Date)
         #years = as.character(years)
         ) %>% 
  #filter(years == '2023' | years == '2022') %>% 
  #group_by(Family.Violence, Highest.Offense.Description, years) %>% 
  group_by(Highest.Offense.Description, years) %>% 
  summarize(cnt = n(), .groups = 'drop') %>% 
  #group_by(Highest.Offense.Description, years) %>% 
  #mutate(tot = sum(cnt)) %>% 
  #relocate(tot, .after = years) %>% 
  mutate(Highest.Offense.Description = as.factor(Highest.Offense.Description)) %>%
  as.data.frame() %>%
  arrange(years) %>% 
  pivot_wider(names_from = years, values_from = cnt)

```

Similar to our zip code variable here, but there are far too many crime descriptions to use. Grouping some together may work, I don't think we have enough information in the dataset overall to do any sort of classification modeling however, so it may amount to a rules based classification using regular expressions. Other variables may still help with this, such as the UCF classification variables.

```{r}
d1 %>% 
  mutate(
    Category.Description = if_else(
      Category.Description == '', 'NA/Unknown', Category.Description
      )
    ) %>% 
  group_by(Category.Description, Highest.Offense.Description) %>% 
  summarize(
    cnt = n()
    ) %>% 
  pivot_wider(
    names_from = Category.Description, values_from = cnt
    )
```

The UCF classifiers are much more consumable, but there are more incidents that are not classified than are. That said, this might be a good start for a rules based classification. Many of the unclassified incidents would not fit very nicely into the groups anyway, so some new categories may need to be made. I may or may not come back to the re-classification part, but ultimately the genesis of this project was with reports on the amount of homicide in the city, so the already existing descriptions may be enough to drill down into just those.

*reminder to myself to save figures from this file to bring into manuscript and presentation later.*

# Data exploration through figures

Time is a variable that is much easier to interpret with a plot, so let's start there.
Time series first.

```{r}
d1 %>% group_by(Occurred.Date) %>% 
  summarize(
  cnt = n()
) %>% as.data.frame() %>% 
  ggplot(mapping = aes(x=Occurred.Date, y=cnt)) +
  geom_line()
```

I expected aggregating by day to actually look worse than this. There is a clear trend, upwards at first but then starts decreasing around 2008 or so. Seeing this I am tempted to looking into time series modeling. There may be a seasonal pattern in some of those spikes, but that may be something that can be resolved with different methods. I'm tempted to go ARIMA, because frankly that's the one I'm familiar with, but I'll have to do some reading up on my options. An interesting idea might be to fit the model, make predictions, and then see where I landed compared to the actual data before the class ends.

```{r}
d1 %>% filter(Occurred.Date <= '2023-12-31') %>%
  mutate(Occurred.Date = trunc.Date(Occurred.Date, 'months')) %>% 
  group_by(Occurred.Date) %>% 
  summarize(
  cnt = n()
) %>% as.data.frame() %>% 
  ggplot(mapping = aes(x=Occurred.Date, y=cnt)) +
  geom_line()
```

Monthly looks even better, again further lending itself to the idea of a time series model. The downward spikes are more prevalent here and may cause problems themselves as well, but again the seasonality may be able to be resolved with things like logs or differencing. As a note, in this and the last plot I filter out 2024, mostly due to a large drop at the end from the incomplete month, but also for the aforementioned idea of comparing predictions to actual results.

A potentially quick way to identify "peak" months for crime would be via a bar chart aggregating all occurrences in each month together. I was hoping to see some quick trends here, but I see surprisingly little here. The highest month is May, potentially because of school letting out/ graduations for UT Austin, but I wouldnt say the difference is large enough compared to other months to really point to any one reason. This may be another one to layer in with time, like a chart with a slider for the year to see shifts in crime frequency by month over time.

```{r}
d2 <- d1 %>% 
  mutate(
    Occurred.Month = month(Occurred.Date, label = TRUE),
    Occurred.Day = weekdays(Occurred.Date),
    Occurred.Year = year(Occurred.Date),
    Occur.Report.Diff = Report.Date - Occurred.Date
  )
d2 %>% 
  #filter(Occurred.Year == 2023) %>%
  ggplot(aes(x=Occurred.Month)) +
  geom_bar()
```

Now a Zip Code bar chart. Really a bit of a graphical representation of what we saw earlier, though I filtered it to the two most recent complete years. Again, looking for big swings, and simultaneously taking a quick peak at the Family Violence indicator. Surprisingly low volume there, which leads me to believe it just doesn't get captured appropriately every time, especially when comparing with the descriptions themselves, which we will see in a moment. Going back to the zip codes, there might be an opportunity to do something like binning the zip codes with high medium and low crime areas, or something of the sort.

```{r}
d1 %>% 
  mutate(years = year(Occurred.Date),
         years = as.character(years)
         ) %>% 
  filter(years == '2023' | years == '2022') %>% 
  group_by(Family.Violence, Zip.Code, years) %>% 
  summarize(cnt = n(), .groups = 'drop') %>% 
  group_by(Zip.Code, years) %>% 
  mutate(tot = sum(cnt)) %>% 
  relocate(tot, .after = years) %>% 
  mutate(Zip.Code = as.character(Zip.Code)) %>%
  as.data.frame() %>% 
  arrange(desc(tot)) %>% head(100) %>% 
  ggplot(aes(fill=Family.Violence, x= reorder(Zip.Code,cnt), y=cnt)) +
  geom_bar(position = 'stack', stat='identity') +
  facet_wrap(~years) +
  coord_flip()
```

Now a bar chart by crime description. This is similar to the last plot, but a little more disparate here. There are so many more crime descriptions than zip codes, so they are much more easily spread thin. That said, that actually makes the top five crime descriptions stand out a little more. So going back to the idea of creating new classifications, that may be a good idea to make sure those 5 are appropriately captured. And again, the family violence indicator is captured almost entirely by one description, yet the top description is Family disturbance. An easy explanation may be because the description is of the Highest Offense in the incident, so when there is physical family violence, it is often the highest offense.

```{r}
d1 %>% 
  mutate(years = year(Occurred.Date),
         years = as.character(years)
         ) %>% 
  filter(years == '2023' | years == '2022') %>% 
  group_by(Family.Violence, Highest.Offense.Description, years) %>% 
  summarize(cnt = n(), .groups = 'drop') %>% 
  group_by(Highest.Offense.Description, years) %>% 
  mutate(tot = sum(cnt)) %>% 
  relocate(tot, .after = years) %>% 
  mutate(Highest.Offense.Description = as.factor(Highest.Offense.Description)) %>%
  as.data.frame() %>% 
  arrange(desc(tot)) %>% head(50) %>% 
  ggplot(aes(fill=Family.Violence, x= reorder(Highest.Offense.Description,cnt), y=cnt)) +
  geom_bar(position = 'stack', stat='identity') +
  facet_wrap(~years) +
  coord_flip() 
```

# Notes

Including this sandbox chunk where I can experiment with code as needed.
ARIMA is not going well, mostly due to seasonal nature of data. Filtering the time range might help, or isolating other variables to do things liek predict certain types of crime instead.

```{r sandbox}

d1 %>% mutate(Category.Description = if_else(Category.Description == '', 'NA/Unknown', Category.Description)) %>% group_by(Category.Description, Highest.Offense.Description) %>% summarize(cnt = n()) %>% pivot_wider(names_from = Category.Description, values_from = cnt)

temp <- d1 %>%
         mutate(Occurred.Date = trunc.Date(Occurred.Date, 'months')) %>% 
         group_by(Occurred.Date) %>% 
         summarize(
             cnt = n()
         ) %>% as.data.frame() %>% 
  filter(year(Occurred.Date) != 2024)
rownames(temp) <- temp$Occurred.Date
temp <- temp %>% select(cnt)

temp2 <- temp %>% as.ts()
Acf(temp2)
Box.test(temp2, lag = 12, type = 'Ljung-Box')
auto.arima(temp2)
temp2 %>% log() %>% diff(12) %>% diff() %>% diff() %>% diff() %>% diff() %>% diff() %>% Acf()
```

### 2024-06-22 code/processing-code/processingfile.qmd
Lots of exploratory work done, deep dive into time series data, and many of the categorical variables in eda code. Moved a lot of that work out of the manuscript. Added descriptions, ideas, and interpretations. Started working on an ARIMA model for crime frequency.

---
title: "An example cleaning script"
author: "Andreas Handel"
date: "2024-02-07"
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

@@

We have six time related fields in this dataset: Occurred.Date.Time, Occurred.Date, Occurred.Time, Report.Date.Time, Report.Date, and Report.Time.\
- The Date-Time and Date variables were read in as string, and the two Time variables were read in as numeric.
So that we can treat these variables as actual representations of a point in time, we will need to re-code them into date, time, and date-time values.\
- Many of the variables have missing or empty values where they shouldn't.
For example, there are some instances where the Occurred.Date.Time variable is empty, but the Occurred.Date and Occurred.Time values are not.\
- Finally, there are some instances where the time value does not make any sense. Values like 60 or 70 are impossible to occur and somewhat vague, so special consideration should be taken to account for these as well.

```{r}
d1 <- rawdata %>% mutate(
  test = (str_sub(as.character(Occurred.Time), -2, -1)
          %>% as.numeric()),
  Occurred.Date = mdy(Occurred.Date)
  ,Report.Date = mdy(Report.Date)
  ,Occurred.Date.Time = parse_date_time(Occurred.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ,Report.Date.Time = parse_date_time(Report.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ) %>% mutate(
    Occurred.Time = if_else(
      is.na(Occurred.Time) | test >= 60, 
      if_else(
        Occurred.Date == Report.Date,
        if_else(
          is.na(Report.Time), 
          2359, Report.Time
          ), 2359
        ), if_else(
          Occurred.Time == 2400, 
          2359, Occurred.Time
          ) 
      ) %>% as.character()
    ) %>% mutate(
      Occurred.Time = paste0('000',Occurred.Time,'00') %>% 
        str_sub(.,-6,-1) %>% 
        parse_time('%H%M%S')
      ) %>% select(-test)
```

Now we see that there is one person with a height of 6.
That could be a typo, or someone mistakenly entered their height in feet.
Since we unfortunately don't know, we might need to remove this person, which we'll do here.

```{r}
d2 <- d1 %>% dplyr::mutate( Height = replace(Height, Height=="6",round(6*30.48,0)) )
skimr::skim(d2)
```

Height values seem ok now.

Now let's look at the `Weight` variable.
There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

```{r}
d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
skimr::skim(d3)
```

Now checking the `Gender` variable.
Gender should be a categorical/factor variable but is loaded as character.
We can fix that with simple base R code to mix things up.

```{r}
d3$Gender <- as.factor(d3$Gender)  
skimr::skim(d3)
```

Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
Well proceed here by removing that individual with that NA entry.
Since this keeps an empty category for Gender, I'm also using droplevels() to get rid of it.

```{r}
d4 <- d3 %>% dplyr::filter( !(Gender %in% c("NA","N")) ) %>% droplevels()
skimr::skim(d4)
```

@@

```{r}
processeddata <- d4
```

@@

Finally, we save the clean data as RDS file.
I suggest you save your processed and cleaned data as RDS or RDA/Rdata files.
This preserves coding like factors, characters, numeric, etc.
If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text.
If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data: http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

# Notes

Removing anyone who had "faulty" or missing data is one approach.
It's often not the best.
based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

```{r Sandbox}
#rawdata %>% filter(Occurred.Date.Time == '')

d1 <- rawdata %>% mutate(
  test = (str_sub(as.character(Occurred.Time), -2, -1)
          %>% as.numeric()),
  Occurred.Date = mdy(Occurred.Date)
  ,Report.Date = mdy(Report.Date)
  ,Occurred.Date.Time = parse_date_time(Occurred.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ,Report.Date.Time = parse_date_time(Report.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ) %>% mutate(
    Occurred.Time = if_else(
      is.na(Occurred.Time) | test >= 60, 
      if_else(
        Occurred.Date == Report.Date,
        if_else(
          is.na(Report.Time), 
          2359, Report.Time
          ), 2359
        ), if_else(
          Occurred.Time == 2400, 
          2359, Occurred.Time
          ) 
      ) %>% as.character()
    ) %>% mutate(
      Occurred.Time = paste0('000',Occurred.Time,'00') %>% 
        str_sub(.,-6,-1) %>% 
        parse_time('%H%M%S')
      ) %>% select(-test)
```

>>>>

---
title: "Cleaning up Crime"
author: "William Hyltin"
date: today
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

@@

We have seven time related fields in this dataset: Occurred.Date.Time, Occurred.Date, Occurred.Time, Report.Date.Time, Report.Date, Report.Time, and Clearance.Date.\
- The Date-Time and Date variables were read in as strings, and the two Time variables were read in as numeric. So that we can treat these variables as actual representations of a point in time, we will need to re-code them into date, time, and date-time values.\
- Many of the variables have missing or empty values where they shouldn't. For example, there are some instances where the Occurred.Date.Time variable is empty, but the Occurred.Date and Occurred.Time values are not.\
- Finally, there are some instances where the time value does not make any sense. Values like 60 or 70 are impossible to occur and somewhat vague, so special consideration should be taken to account for these as well.

```{r}
d1 <- rawdata %>% mutate(
  test = (str_sub(as.character(Occurred.Time), -2, -1)
          %>% as.numeric()),
  Occurred.Date = mdy(Occurred.Date)
  ,Report.Date = mdy(Report.Date)
  ,Occurred.Date.Time = parse_date_time(Occurred.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ,Report.Date.Time = parse_date_time(Report.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ,Clearance.Date = mdy(Clearance.Date)
  ) %>% mutate(
    Occurred.Time = if_else(
      is.na(Occurred.Time) | test >= 60, 
      if_else(
        Occurred.Date == Report.Date,
        if_else(
          is.na(Report.Time), 
          2359, Report.Time
          ), 2359
        ), if_else(
          Occurred.Time == 2400, 
          2359, Occurred.Time
          ) 
      ) %>% as.character()
    ) %>% mutate(
      Occurred.Time = paste0('000',Occurred.Time,'00') %>% 
        str_sub(.,-6,-1) %>% 
        parse_time('%H%M%S')
      ,Occurred.Date.Time = if_else(
        is.na(Occurred.Date.Time), 
        ymd_hms(paste(Occurred.Date, Occurred.Time)), Occurred.Date.Time
        )
      ,Report.Time = if_else(
        is.na(Report.Time), 2359, Report.Time) %>%
        as.character() %>% 
        paste0('000',.,'00') %>% 
        str_sub(.,-6,-1) %>% 
        parse_time('%H%M%S')
      ,Report.Date.Time = if_else(
        is.na(Report.Date.Time),
        ymd_hms(paste(Report.Date, Report.Time)), Report.Date.Time
      )
      ) %>% select(-test)
skim(d1)
```

This should address all of the time-based variables. I may come back and update some of this, solely because I made some more analytic decisions regarding the empty values for Occurred.Date.Time. The logic is that if the time is missing, the occurrence would have had to have happened at least before the report time, though if the report occurred on a different day we assume the occurrence would have to be at least by midnight the same day. The number of empties, once the three "Occurred" variables are coalesced where appropriate, are relatively low, so I don't expect this to have significant effect on any results either way.

### Categorical Variables

In the initial looks at the dataset a few columns stand out as potential factor variables.  
 - Family Violence is a mostly complete field, though a minor oddity to it is the existence of a third factor when the codebook says it should only be two.  
 - APD Sector, APD District, and PRA may be the best values to serve as a categorical representation of location, since it has relatively few missing variables. Zip code may also work and be more consumable to the average person, however it does have more missing than the others mentioned. Lastly, X and Y Coordinates, though not naturally categorical, may be able to be used to create categorical groupings considering their completeness.  
 - Location Type is another good categorical variable. There are a lot of categories, but with one like this it may be easy to combine some of them. Empty values are climbing, but still less than 1% of the data.  
 - Highest Offense code has 396 distinct values, which would be far too many for categorical use, but again may be able to be grouped.
 - Clearance Status sounds like a very interesting variable to use, however it is one of the variables with the highest number of empties/missing. Furthermore, it's confounded by Clearance Date, which has fewer (but still a lot of) missing and empty values. This is one that may take some outside research, on what constitutes a "clearance" and what it means in relation to a clearance date.
 - UCR Category and Description also sound like very interesting variables, but are the worst offenders in terms of missing data. This makes sense from info from the codebook, that it is reserved for designation of the most serious crimes by the FBI. In a way this could even mean the null values are also valuable, because they very well may indicate less serious crimes.

```{r}
d2<- d1 %>% 
  mutate(
    Family.Violence = toupper(Family.Violence)
    ,Location.Type = if_else(
      Location.Type == '',
      'Unknown', Location.Type
      )
    ) %>% 
  filter(!is.na(Zip.Code) & !(Zip.Code == 0) & !is.na(X.coordinate) & !is.na(Y.coordinate) & PRA != '',Clearance.Status != 9)
skim(d2)
print(count(d2)/count(d1))
```

The categorical variables look good for now. Up until this point little exploration has been done and there may be other re-codes or filters that would be worth including for other variables. Likely once I am further into the analysis I will return to this step to make more adjustments, but the changes I've made should provide a good starting point.

### String Variables

There are a few string variables that may be worth using for the analysis. Some of these I covered in the categorical variables, but for those that have excessive number of unique values a categorical value may not be practical.  
 - Highest Offense Description has hundreds of unique values, but since this is essentially a description of the crimes committed there are likely commonalities that could lead to groupings. I can imagine more variables stemming from this as well, such as if the description contains certain keywords I could create indicator variables based on that presence.
 - Address is essentially another location based variable, and considering the number of them that have been cleaned for that purpose already it seems unlikely to need more. Still, the specificity offered by the address may be insightful, especially if i start drilling down into particular areas that suffer from the most crime. Potentially Doing something like a regular expression to extract a street name would be useful.  
 - Location Type does not have an unreasonable number of unique values, so this variable may better fit as a categorical only. Still the max length of some of the values lends itself to be that there may be more worth extracting from the field.

Parsing out these string variables will likely be more of a cognitive load than the others would, so I will hold off on doing so for now. Once I am further along with the EDA I will have a better idea of what I want to be looking for within these variables anyway.

### Tidied Up

@@

```{r}
processeddata <- d2
```

@@

Finally, we save the clean data as RDS file.

```{r}
save_data_location <- here("data","processed-data","processed-crime.rds")
saveRDS(processeddata, file = save_data_location)
```

```{r Sandbox, echo=FALSE}
#rawdata %>% filter(Occurred.Date.Time == '')

rawdata %>% mutate(
  test = (str_sub(as.character(Occurred.Time), -2, -1)
          %>% as.numeric()),
  Occurred.Date = mdy(Occurred.Date)
  ,Report.Date = mdy(Report.Date)
  ,Occurred.Date.Time = parse_date_time(Occurred.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ,Report.Date.Time = parse_date_time(Report.Date.Time,'%m/%d/%Y %I:%M:%S %p')
  ,Clearance.Date = mdy(Clearance.Date)
  ) %>% mutate(
    Occurred.Time = if_else(
      is.na(Occurred.Time) | test >= 60, 
      if_else(
        Occurred.Date == Report.Date,
        if_else(
          is.na(Report.Time), 
          2359, Report.Time
          ), 2359
        ), if_else(
          Occurred.Time == 2400, 
          2359, Occurred.Time
          ) 
      ) %>% as.character()
    ) %>% mutate(
      Occurred.Time = paste0('000',Occurred.Time,'00') %>% 
        str_sub(.,-6,-1) %>% 
        parse_time('%H%M%S')
      ) %>% select(-test)
```

### 2024-06-22 products/manuscript/Manuscript.qmd
Minor changes made.
Lots of exploratory work done, deep dive into time series data, and many of the categorical variables in eda code. Moved a lot of that work out of the manuscript. Added descriptions, ideas, and interpretations. Started working on an ARIMA model for crime frequency.

```{r Setup, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

```{r Load Processed Data}
d1 <- readRDS(here('data','processed-data','processedcrime.rds'))
```

@@

>>>>

```{r Setup, echo=FALSE, message=FALSE}
# load a few R packages
pacman::p_load(here, knitr, tidyverse, skimr)
theme_set(theme_minimal())
```

@@

```{r Load Processed Data}
d1 <- readRDS(here('data','processed-data','processedcrime.rds'))
zips <- read_csv(here('data','raw-data','austin-zip-codes.csv'))
```

@@

```{r sandbox}
ggplot() +
    geom_sf(aes(fill = zip.code), data = zips2, inherit.aes = F, size = 0, alpha = 0.6, geom = zips2$the_geom) +
    coord_sf(ndiscr = F) +
    theme(legend.position = "none")
```
